---
title: "Statisical Inference Project"
author: "James Brink"
date: "September 18, 2014"
output: pdf_document
---

##Part 1: Simulation Exercises

In this part we will create a simulation of an exponential distribution with the rexp() function in R.  To do this we will do 1000 simulations of the mean of 40 exponentials using a constant lambda of 0.2 for all of our simulations.

First we will simulate our data and populate a data frame called "meanData" with the means from our simulations.
```{r}
##set variables
set.seed(7)
lambda <- 0.2
n <- 40
sims <- 1000

##simulate means and insert into data frame
meanData <- sapply(1:sims, function(x) {mean(rexp(n, lambda))})
meanData <- as.data.frame(meanData)
names(meanData)[1] <- "x"
head(meanData)
```

Now we can look at some of the calculated attributes of our simulated distribution and compare them to their theorhetical values.
```{r}
##the center of our simulated distribution
simDistMean <- mean(meanData$x)
simDistMean
```

```{r}
##the theorhetical center of our distribution
theoDistMean <- 1/lambda
theoDistMean
```

We see that the distribution that we simulated is centered around `r simDistMean`, very close to the theorhetical center of the distribution which is `r theoDistMean`. 

Next we examine the variability of our simulation distribution and compare it to the theorhetical variablity.
```{r}
##the standard deviation of our simulated distribution
simDistSd <- sd(meanData$x)
simDistSd
```

```{r}
##the theorhetical center of our distribution
theoDistSd <- (1/lambda)/sqrt(40)
theoDistSd
```

```{r}
##the standard deviation of our simulated distribution
simDistVar <- var(meanData$x)
simDistVar
```

```{r}
##the theorhetical center of our distribution
theoDistVar <- theoDistSd^2
theoDistVar
```
Our variability measures are close to our the theoreticals as well.  We see that the distribution that we simulated is has a standard deviation of `r simDistSd` and a variance of `r simDistVar`, while the theorhetical standard deviation of the distribution is `r theoDistSd` and the theorhetical variance is `r theoDistVar`.

Below we examine the simulated data using a histogram and then plot a density line over top of it in red and a density line for a normal distribution over top as well in black. We can see that the simulated exponential distribution is approximately normal.  We also show the simulated distribution mean as a black vertical line.
```{r}
library(ggplot2)
g <- ggplot(meanData, aes(x=x))
g <- g + geom_histogram(aes(y=..density..), fill="#CCFFCC", color="black", binwidth=0.15)
g <- g + stat_density(color="red", geom="line", position="identity")
g <- g + stat_function(fun = dnorm, arg = list(sd = simDistSd, mean = simDistMean))
g <- g + geom_vline(aes(xintercept=theoDistMean))
g <- g + ggtitle("Simulated Distribution")
g
```

Lastly we can evaluate the coverage of +/- 1.96 standard deviations on the simulated distribution and see if it is similar to that of a normal distribution.  We know that +/- 1.96 standard deviations away from the mean provides us with a 95% confidence interval on a normal distribution, lets evaluate the coverage +/-1.96 standard deviations away from our simulated mean.

```{r}
cuts <- theoDistMean + c(-1,1)*1.96*((1/lambda)/sqrt(40))
coverage <- length(meanData[meanData$x > cuts[1] & meanData$x < cuts[2],]) / nrow(meanData)
coverage
```
We see that `r coverage*100`% of our simulated sample means fall within 1.96 standard deviations of the population mean.  This is very close to the 95% we know to be true in a normal distribution.
